Exact Asymptotics for Causal Mediation Analysis

### 2024-05-17 notes/UQ Verification.tex
Gradient of Phi function for bin-bin-mixed case in R

\subsection{UQ for GLMM Parameters}

@@

To this end, I need the Hessian of the observed data (log) likelihood. There are tricks from the EM algorithm literature for doing this, but I think our problem is simple enough that I can just reduce it to a small-ish number of one-dimensional integrals. To start, recall the formula Bruno gives for $\bP(Y=1 | X=x, \mathbf{W}=\mathbf{w}) =: \bP_1$. Specifically, he expresses this quantity in terms of the $\phi$ function evaluated at four different arguments. Thus, if I have the Hessian of $\phi$, I can compute the Hessian of $\bP_1$ (and $\bP_0$), which in-turn gives me the Hessian of the log-likelihood based on a single point.

@@

>>>>

The approximate covariance matrix of the GLMM parameter estimates can be obtained directly from a fitted \texttt{lme4} object using the \texttt{merDeriv} package in \texttt{R}. This saves all the work I was starting in the following subsection. Please disregard.

\subsection{UQ for GLMM Parameters (Old)}

@@

To this end, I need the Hessian of the observed data (log) likelihood. There are tricks from the EM algorithm literature for doing this, but I think our problem is simple enough that I can just reduce it to a small-ish number of one-dimensional integrals. To start, recall the formula Bruno gives for $\bP(Y=1 | X=x, \mathbf{W}=\mathbf{w}) =: P_1$ (similar for $P_0$). Specifically, he expresses this quantity in terms of the $\phi$ function evaluated at four different arguments. Thus, if I have the Hessian of $\phi$, I can compute the Hessian of $P_1$ (and $P_0$), which in-turn gives me the Hessian of the log-likelihood based on a single point.

@@

\subsection{UQ for Mediation Effects}

Following Section 2.2 of B \& B's document, we have an expression for the mediation effect, $\Psi$, in terms of a small number of integrals, $\phi$. Each of these integrals is univariate, and can thus be computed accurately using quadrature. As for uncertainty quantification, we still need to compute the partial derivative of $\Psi$ wrt each of its arguments. This shouldn't be too hard to do in Maple (hopefully). I will probably need to leave $\phi$ undefined, and get Maple to return an answer in terms of the partial derivatives of $\phi$.

Note that $\Psi$ depends explicitly on a combination of model parameters and transformations thereof. In particular, the functions $\sigma_U$ and $\sigma_V$ depend implicitly on the mixed effects' covariance matrices. Conveniently, B \& B have worked out the gradients of $\sigma_U$ and $\sigma_V$ wrt the model parameters. They're messy, but ultimately code-able.

\textcolor{red}{As of Friday, May 17, I have coded the gradient of $\Phi$. Next I need to do UQ for each argument of $\Phi$ (e.g., $\sigma_V(x+1)$), then I can apply the $\delta$-method.}

### 2024-05-17 notes/William Analysis.Rmd
Gradient of Phi function for bin-bin-mixed case in R

We first demonstrate our analysis on a single dataset, then put this analysis inside a `for` loop. The first step is to set the sample size and generate some data. We use $\eta$ to denote the linear predictor for $M$, and $\zeta$ for the linear predictor of $Y$ without the $M$ term. That is, $\zeta = b_0 + b_2 X + W B_3$.

@@

>>>>

We first demonstrate our analysis on a single dataset, then put this analysis inside a `for` loop. The first step is to set the sample size and generate some data. We use $\eta$ to denote the linear predictor for $M$, and $\zeta$ for the linear predictor of $Y$. <!-- (This is a comment) without the $M$ term. That is, $\zeta = b_0 + b_2 X + W B_3$. -->

@@

# Binary Response, Binary Mediator, Mixed-Effects

We now introduce random effects to our model. This changes many of the details, but the overall structure of the analysis remains the same. I still need to write some helper functions to compute relevant derivatives, all of which validated against a simple finite-difference approximation. See `Helpers.R`.

```{r}
source("../src/Helpers.R")

library(lme4)     # For fitting GLMMs using glmer()
library(MASS)     # For simulating multivariate normals using mvrnorm()
library(merDeriv) # For computing the full information matrix from models fit using lme4
```

As in the previous case, we set intercepts so that the means of both linear predictors are at least approximately zero. We also need to set covariance parameters for the random effects. For this, we use either $20\%$ of the true value or $0.2$ if the true value is zero. Correlations between all random effects are set to $0.5$. Note that we include random effects for the intercept and the exposure, but not for the mediator.

Another difference from the previous settings is that we must now construct our dataset in clusters. We will use $K=5$ such clusters, with $n$ observations in each cluster. We will set $n=100$ for the first iteration of the loop.

```{r setup_mixed}
# Fixed effects for intercepts
a_0 = 0
b_0 = -0.5

# Random effects
sigma_a_0 = 0.2
sigma_a_1 = 0.2 * abs(a_1)
cor_a0_a1 = 0.2
cov_a0_a1 = sigma_a_0 * sigma_a_1 * cor_a0_a1
Sigma_a = matrix(c(sigma_a_0^2, cov_a0_a1, cov_a0_a1, sigma_a_1^2), nrow = 2, ncol = 2)

sigma_b_0 = 0.2 * abs(b_0)
sigma_b_1 = 0.2 * abs(b_1)
cor_b0_b1 = 0.2
cov_b0_b1 = sigma_b_0 * sigma_b_1 * cor_b0_b1
Sigma_b = matrix(c(sigma_b_0^2, cov_b0_b1, cov_b0_b1, sigma_b_1^2), nrow = 2, ncol = 2)

# Number of clusters
K=5
```

We first demonstrate our analysis on a single dataset, then put this analysis inside a `for` loop. The first step is to set the sample size and generate some data. We use $\eta$ to denote the linear predictor for $M$, and $\zeta$ for the linear predictor of $Y$. <!-- (This is also a comment) without the $M$ term. That is, $\zeta = b_0 + b_2 X + W B_3$. -->

```{r generate_data_mixed}
n = all_Ns[2]
# n = all_Ns[1]

all_Xs = list()
all_Ws = list()

for(k in 1:K){
  X = rnorm(n, mean=0, sd=1)
  W = matrix(rnorm(n*p_conf, mean=0, sd=1), nrow = n, ncol = p_conf)

  all_Xs[[k]] = X
  all_Ws[[k]] = W
}

# Generate M
all_Ms = list()
for(k in 1:K){
  eta_vec_fixed = a_0 + a_1*all_Xs[[k]] + all_Ws[[k]]%*%A_2

  ## Add random effects
  a_ran = mvrnorm(1, mu = rep(0, 2), Sigma = Sigma_a)
  eta_vec = eta_vec_fixed + a_ran[1] + a_ran[2]*all_Xs[[k]]

  ## Generate M
  p_M_vec = expit(eta_vec)
  M = rbinom(n, size = 1, prob = p_M_vec)
  all_Ms[[k]] = M
}

# Generate Y
all_Ys = list()
for(k in 1:K){
  zeta_vec_fixed = b_0 + b_1*all_Ms[[k]] + b_2 * all_Xs[[k]] + all_Ws[[k]]%*%B_3

  ## Add random effects
  b_ran = mvrnorm(1, mu = rep(0, 2), Sigma = Sigma_b)
  zeta_vec = zeta_vec_fixed + b_ran[1] + b_ran[2]*all_Xs[[k]]

  ## Generate Y
  p_Y_vec = expit(zeta_vec)
  Y = rbinom(n, size = 1, prob = p_Y_vec)
  all_Ys[[k]] = Y
}

# Consolidate groups
X = do.call(c, all_Xs)
W = do.call(rbind, all_Ws)
M = do.call(c, all_Ms)
Y = do.call(c, all_Ys)
group = rep(1:K, each = n)
```

```{r fit_models_mixed}
M_data = data.frame(M, X, W1 = W[, 1], W2 = W[, 2], W3 = W[,3], group = group)
M_model = glmer(M ~ X + W1 + W2 + W3 + (1 + X | group), data = M_data, family = binomial(link = "logit"))
M_model_info = attributes(VarCorr(M_model)$group)

a_hat = fixef(M_model)
a_RE_sds = M_model_info$stddev
a_RE_cor = M_model_info$correlation[2,1]
a_cov = vcov(M_model, full=TRUE, ranpar="sd")

# a_hat = summary(M_model)$coefficients[, 1]
# a_SE = summary(M_model)$coefficients[, 2]
# a_cov = vcov(M_model)

Y_data = data.frame(Y, M, X, W1 = W[, 1], W2 = W[, 2], W3 = W[,3], group = group)
Y_model = glmer(Y ~ M + X + W1 + W2 + W3 + (1 + X | group), data = Y_data, family = binomial(link = "logit"))
Y_model_info = attributes(VarCorr(Y_model)$group)

b_hat = fixef(Y_model)
b_RE_sds = Y_model_info$stddev
b_RE_cor = Y_model_info$correlation[2,1]
b_cov = vcov(Y_model, full=TRUE, ranpar="sd")

# b_hat = summary(Y_model)$coefficients[, 1]
# b_SE = summary(Y_model)$coefficients[, 2]
# b_cov = vcov(Y_model)
```
